---
title: 'Interim Data Analysis : Associations with Adult Orbital Cellulitis'
author: "Ryan Summers"
date: "2026-04-19"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    css: style.css
  bookdown::pdf_document2: default
bibliography: oc_trends.bib
csl: "apa-5th-edition.csl"
nocite: '@*'
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(readxl)
library(splines)
library(MASS)
library(dplyr)
library(gt)
library(naniar)
library(performance)
library(ggcorrplot)
library(gtsummary)
library(ncvreg)
library(StepReg)
library(logistf)
library(glmnet)
library(pROC)

# scripts
source("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/Orbital-Cellulitis-Research/adult_analysis/Code/group_counts_ftn.R")
```

## Introduction

>The data for this analyses consists of hospital admissions for orbital cellulitis 
from January 2020 through September 2025, from a retrospective chart review across 
three hospital systems. Orbital cellulitis admission data were obtained from 
Denver Health (DH), University of Colorado (UC), and Oregon Health and Science (OHSU).
This dataset includes

Viral variables include influenza, respiratory syncytial virus (RSV), adenovirus, human metapneumovirus (HMPV), parainfluenza viruses (types 1–4), rhinovirus/enterovirus, and coronavirus. Respiratory virus surveillance data were obtained from the Centers for Disease Control and Prevention (CDC). 
>
The primary objectives of this analysis were:
>
>1.	To evaluate whether monthly viral detection counts is associated with orbital cellulitis admissions.
2.	To assess correlations among viruses to determine whether multicollinearity may affect model stability.
3.	To determine whether these associations occur with delayed effects across 0–3-month lag periods.
4.	To compare alternative DLNM specifications and identify the most appropriate lag–exposure structure.

1. What are the factors influencing need for surgery? 
    - age
    - sinusitis
    - diabetes
    - immunocompromise
    - drugs
    - asthma
    - trauma
    - prior surgery
    - dental 
    - NLDO
    - SPA
    - OA
    - ICE
2. What are the factors influencing length of stay (same as before) 
3. What are the factors influencing development of SPA? OA? ICE? 

## Methods

### Data Management & Cleanup

```{r, warning=FALSE, message=FALSE}
oc_raw <- read_excel("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/Orbital-Cellulitis-Research/adult_analysis/DataRaw/OHSU\ and\ UC.xlsx", sheet="Adults")
```

A priori variables based on discussion with PI
```{r}
priori <- c("age", "sinusitis", "diabetes", "immunocompromise", "drug/alc",
            "asthma", "trauma", "prior_surgery", "surgery_adm",
            "dental_infection",  "nldo/infection", "spa", "oa", "ice")
```


```{r,warning=FALSE, message=FALSE}
# cleanup column names
oc_cleaned <- oc_raw %>% 
  # lowercase columns & rows & remove white-space in columns
    setNames(names(.) |> tolower() |> gsub("\\s+", "_", x = _) |> 
               make.unique(sep = "_dup_")) %>%
  # lower case and remove white-space from rows
  mutate(across(where(is.character), \(x) str_squish(tolower(x)))) %>%
  # for stability mark other sex as NA since its only one case
  mutate(sex = ifelse(sex == "other", NA, sex),
         # re-calculate age variable (og is AAA)
         age = as.numeric(difftime(doa, dob, units = "days") / 365.25),
         sex = factor(sex, levels = c(0,1), labels = c("Female", "Male"))) %>% 
  # relocate age
  relocate(age, .after = doa) %>% 
  # rename(oc_adm = orbital_cellulitis_admissions,
  #        rv_ev = `rv/ev`) %>% 
  # remove last two rows containing summaries
  filter(!is.na(record_id)) %>% 
  # rename duplicate surgery variables
  rename(
    prior_surgery = surgery...28,
    surgery_adm = surgery...61) %>% 
  # convert variables to factors
  mutate(across(all_of(priori[priori != "age"]),
                ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes"))),
         # collapse race/ethnicity into one column vector
         race_eth = case_when(
           hl == 1 ~ "Hispanic",
           nonhl == 1 & white == 1 ~ "White non-Hispanic",
           nonhl == 1 & black == 1 ~ "Black non-Hispanic",
           nonhl == 1 & `asian/pi`== 1 & native_american == 1 & other...10 == 1
           ~ "Other",
           TRUE ~ NA_character_))

# duplicate names
# which(names(oc_cleaned) == "surgery")
# names(oc_cleaned)[grepl("surgery", names(oc_cleaned))]

```


```{r}
# find duplicated column names
names(oc_raw)[duplicated(names(oc_raw))]
which(duplicated(names(oc_raw)))

# check where in my pipeline its creating a duplicate
n1 <- names(oc_raw)
n2 <- tolower(n1)
n3 <- gsub("\\s+", "_", n2)

n1[duplicated(n1)]
n2[duplicated(n2)]
n3[duplicated(n3)]
```
> The race/ethnicity covariate was categorized as White non-Hispanic, Black non-Hispanic, and Hispanic. We acknowledge that race is a social construct and that observed differences likely reflect structural inequities rather than biological differences.  

### Exploratory Data Analysis (EDA)


```{r label = "Used in final report as Table 1"}
tbl1 <- oc_cleaned %>%
  tbl_summary(
    by = surgery_adm,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, spa, oa, ice),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} ({p25}, {p75})",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    #igits = list(AGE ~ 0, SYSBP ~ 0),
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromise",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection`~ "NLDO",
      spa ~ "SPA",
      oa ~ "OA",
      ice ~ "ICE"
      )) %>%
  add_overall() %>% bold_labels() %>% italicize_levels() %>% add_n() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Surgery Needed**") %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of Orbital Cellulitis Adult Cohort") %>% 
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```


```{r}
# change label names for hospitals
unique(oc_cleaned$hospital)
```

Number of patients
```{r}
cat("Number of patients:",
  length(unique(oc_cleaned$record_id)),"\n\n")
```

Missing values
```{r}
oc_cleaned %>% ungroup() %>% miss_var_summary()
# missing values for predictors of interest
oc_cleaned[, priori] %>% ungroup() %>% miss_var_summary()
```


**NLDO** = Nasolacrimal duct infection:
a tear duct inflammation and/or infection, can happen in newborn infants and in
adults. Treatment usually includes antibiotics. Surgery is a possible option to 
stop recurrences.

**Dacryocystitis** is an infection of the tear sac (lacrimal sac) caused by a 
blockage in the tear duct (nasolacrimal duct), leading to pain, redness, and 
swelling around the eye. Treatment typically involves antibiotics, and in some 
cases, surgery may be necessary to relieve the blockage.

Neutrophils = innate, acute inflammation
Lymphocytes (T,B,NK) = adaptive, regulated immunity
  * Neutrophils ↑ = acute inflammation, stress, infection
  * Lymphocytes ↓ = stress response, immune suppression, chronic disease
  * So high NLR = More inflammation + relatively weaker adaptive immune response
  
Physiological meaning of high vs low NLR

High NLR:
  * Acute stress or infection
  * Chronic inflammation
  * Poor prognosis in some diseases

Low NLR:
  * Lower inflammation
  * Stronger relative adaptive immunity (context-dependent)

Get group counts
```{r}
count_grp_obs(oc_cleaned, vars = priori[priori != "age"], 
              mode = "separate")
```

```{r}
sapply(oc_cleaned[priori[priori != "age"]], table) 
```
Correlations
```{r label = "Correlations"}
# need to convert back to numeric
X_bin <- oc_cleaned %>%
  dplyr::select(all_of(priori[priori != "age"])) %>%
  mutate(across(everything(), ~ as.numeric(.x) - 1))

clean_names <- c(
  "Sinusitis", "Diabetes", "Immunocompromise", "Drug/Alc",
  "Asthma", "Trauma", "Prior Surgery", "Surgery Needed",
  "Dental Infection", "NLDO Infection", "SPA", "OA", "ICE")

cor_matrix <- cor(X_bin, use = "pairwise.complete.obs", method = "pearson")
colnames(cor_matrix) <- clean_names
rownames(cor_matrix) <- clean_names

ggcorrplot(cor_matrix, type = "lower",lab = T)
```



Per Frank Harrell - Events per varialbe (EPV) ≈ 10 EPV rule
$EPV = \frac{\text{Events}}{\text{Predictors}}$

Variable Selection

```{r}
form <- surgery_adm ~ .

step.fit <- stepwise(formula = form,
                data = oc_cleaned,
                type = "logit",
                include = c("age", "sex", "race_eth", "sinusitis", "diabetes",
                            "immunocompromise", "drug/alc", "asthma", "trauma", 
                            "prior_surgery", "dental_infection", "nldo/infection", 
                            "spa", "oa", "ice"),
                strategy = "backward",
                metric = c("AICc", "SBC"))

summary(step.fit$backward$AICc)
summary(step.fit$backward$SBC)
performance(step.fit)
```
```{r}
# Check unique values and class of each predictor
vars_to_check <- c("age", "sex", "race_eth", "sinusitis", "diabetes",
                   "immunocompromise", "drug/alc", "asthma", "trauma", 
                   "prior_surgery", "dental_infection", "nldo/infection", 
                   "spa", "oa", "ice")

for (v in vars_to_check) {
  if (v %in% names(oc_cleaned)) {
    cat(v, ":", 
        "class =", class(oc_cleaned[[v]]), 
        "unique values =", length(unique(na.omit(oc_cleaned[[v]]))), 
        "\n")
  } else {
    cat(v, "NOT FOUND in data\n")
  }
}
```

### Primary Analysis

**logistf**
logistic regression model using Firth's bias reduction method, equivalent to 
penalization of the log-likelihood by the Jeffreys prior
  * Profile likelihood = method used in statistics to estimate the likelihood 
  of a parameter by fixing other parameters and optimizing the remaining ones
  * Penalized likelihood = method that incorporates a penalty term into the 
  likelihood function to prevent overfitting and improve model estimation, 
  especially in cases with complex models or limited data.
  
> Variables were selected using penalized regression and then entered into a 
Firth logistic model for estimation.

> Candidate predictors were screened using cross-validated penalized logistic 
regression. Variables selected by glmnet were then entered into a Firth penalized 
logistic regression model along with prespecified adjustment covariates: age, 
sex, and race/ethnicity.
  
```{r}
vars_quoted <- ifelse(grepl("/", vars_to_check), 
                      paste0("`", vars_to_check, "`"), vars_to_check)
form <- as.formula(paste("surgery_adm ~", paste(vars_quoted, collapse = " + ")))
nfs.model.fbm <- logistf(form,data=oc_cleaned)

summary(nfs.model.fbm)
```

GLM
```{r}
nfs.model.glm <- glm(form,family="binomial",data=oc_cleaned)

summary(nfs.model.glm)
```

GLMnet
```{r}
# handles NA removal to be consistent with dimensions of outcome
mf <- model.frame(form, data = oc_cleaned)

X <- model.matrix(form, data = mf)[, -1] # remove y vector
y <- mf$surgery_adm
```

```{r}
# set so that covariates have a penalty factor of 0 so theyre always retained
penalty_factor <- rep(1, ncol(X))
penalty_factor[
  colnames(X) %in% c("age", "sexMale", "race_ethWhite non-Hispanic", 
                     "race_ethHispanic")] <- 0
cvfit <- cv.glmnet(
  X, y,
  family = "binomial",
  alpha = 1,
  type.measure = "class",
  penalty.factor = penalty_factor)

plot(cvfit)
```

extract best model
```{r}
coef(cvfit, s = "lambda.min")
coef(cvfit, s = "lambda.1se")
# variables selected per cross-validation results
rownames(coef(cvfit, s = "lambda.1se"))[
  as.numeric(coef(cvfit, s = "lambda.1se")) != 0]
```

Predictions and model performance
```{r}
pred_prob <- predict(cvfit, newx = X, s = "lambda.min", type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

# ROC curve - evaluate PC1 as a predictor of malignancy on test set
roc_obj = roc(y, pred_prob)

# Find optimal sensitivity, specificity, and threshold using Youden's method
opt_coords <- coords(roc_obj, "best", 
                     ret = c("threshold", "sensitivity", "specificity"), 
                     best.method = "youden")

# Plot the ROC curve and display the AUC
plot(roc_obj, print.auc = TRUE, xlim = c(1, 0), ylim = c(0, 1))
text(0.4, 0.20, paste("Optimal Threshold:", round(opt_coords$threshold, 3)), 
     cex = 0.9, col = "steelblue")
text(0.4, 0.15, paste("Sensitivity:", round(opt_coords$sensitivity, 3)), 
     cex = 0.9, col = "steelblue")
text(0.4, 0.10, paste("Specificity:", round(opt_coords$specificity, 3)), 
     cex = 0.9, col = "steelblue")

auc(roc_obj)
```


#### Final Model

```{r}
# OR & 95% CI
fbm_tbl <- data.frame(
  #term = names(nfs.model.fbm$coefficients),
  OR = exp(nfs.model.fbm$coefficients),
  LL = exp(nfs.model.fbm$ci.lower),
  UL = exp(nfs.model.fbm$ci.upper),
  p_value = nfs.model.fbm$prob) %>% 
  mutate(
    # % change in odds (per unit increase)
    pct_chg  = (OR  - 1) * 100,
    pct_LL = (LL - 1) * 100,
    pct_UL = (UL - 1) * 100) %>% 
  mutate(pct_ci  = sprintf("(%.2f, %.2f)", pct_LL, pct_UL)) 

fbm_tbl
```


```{r label = "Used in final report as table XX"}
nfs_tbl <- fbm_tbl %>%
  mutate(n_fbm= nrow(model.frame(nfs.model.fbm))) %>% 
  select(n_fbm, OR, pct_chg, pct_ci, p_value) %>%
  gt(rowname_col = "Predictor") %>%
  cols_label(
    n_fbm = "N", 
    OR = "OR",
    pct_chg = "%Δ in Odds",
    pct_ci = "95% HDI",
    p_value  = "P-value") %>%
  # tab_spanner(
  #   label = "Bayesian",
  #   columns = c(n_brm, bay_pct, bay_ci)) %>%
  fmt_number(columns = pct_ch, decimals = 2) %>%
  fmt_number(columns = p_value, decimals = 3) %>%
    # bold specific text
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  # tab_style(
  #   style = cell_text(weight = "bold"),
  #   locations = cells_column_spanners(
  #     spanners = c("Frequentist", "Bayesian"))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = freq_p),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "Values represent percent change in the odds of needing surgery
    per one-unit increase in the predictor (computed as (OR − 1) × 100).
    Negative values indicate lower odds of surgical need.") %>% 
  tab_options(
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl_dropout_compare
```




