---
title: 'Interim Data Analysis : Associations with Adult Orbital Cellulitis'
author: "Ryan Summers"
date: "2026-04-19"
output: 
  word_document: default
  bookdown::pdf_document2: default
  bookdown::html_document2:
    toc: true
    toc_float: true
    code_folding: hide
    css: style.css
bibliography: oc_adults.bib
csl: "apa-5th-edition.csl"
nocite: '@*'
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(tidyverse)
library(magrittr)
library(ggfortify)
library(olsrr)
library(readxl)
library(splines)
library(MASS)
library(dplyr)
library(gt)
library(naniar)
library(performance)
library(ggcorrplot)
library(gtsummary)
library(ncvreg)
library(StepReg)
library(logistf)
library(glmnet)
library(pROC)

library(flextable)

# scripts
source("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/Orbital-Cellulitis-Research/adult_analysis/Code/group_counts_ftn.R")
```

## Introduction

> Orbital cellulitis (OC) is a serious infection of the orbital tissues that can lead to vision-threatening and life-threatening complications, often requiring hospitalization and, in severe cases, surgical intervention. Clinical presentations can vary, ranging from localized inflammation to the development of orbital and intracranial complications such as subperiosteal abscess (SPA), orbital abscess (OA), and intracranial extension (ICE). 
>
> The data for these analyses consist of hospital admissions for OC from June 2006 through May 2025, derived from a retrospective chart review across three hospital systems: Denver Health (DH), University of Colorado (UC), and Oregon Health and Science University (OHSU). The dataset includes 330 adult subjects with detailed clinical, demographic, and outcome information collected at the time of admission.
>
> The primary objectives of this analysis were to identify factors associated with need for surgical intervention and length of hospital stay (LOS) among adults admitted for orbital cellulitis. Secondary analyses examined factors associated with the development of complications, including SPA, OA, and ICE. We hypothesized that markers of disease severity and underlying comorbidities would be associated with increased likelihood of surgery and prolonged LOS.


## Methods

* OR - relates to probability of an event
* RR - relates to amount of something (counts/time)
  * OR=2 ≠ RR=2 - OR is usually more extreme when event is more common
  
Logistic regression models estimated odds ratios (ORs) for binary outcomes, 
representing the relative odds of an event. Negative binomial models estimated 
rate ratios (RRs) for length of stay, representing multiplicative changes in 
expected LOS.

### Data Management & Cleanup

```{r, warning=FALSE, message=FALSE}
oc_raw <- read_excel("/Users/ryan_summers/Library/Mobile\ Documents/com~apple~CloudDocs/GitHub/Orbital-Cellulitis-Research/adult_analysis/DataRaw/OHSU\ and\ UC.xlsx", sheet="Adults")
```

A priori variables based on discussion with PI
```{r}
priori <- c("age", "sinusitis", "diabetes", "immunocompromise", "drug/alc",
            "asthma", "trauma", "prior_surgery", "surgery_adm",
            "dental_infection",  "nldo/infection", "spa", "oa", "ice")
```


```{r,warning=FALSE, message=FALSE}
# cleanup column names
oc_cleaned <- oc_raw %>% 
  # lowercase columns & rows & remove white-space in columns
    setNames(names(.) |> tolower() |> gsub("\\s+", "_", x = _) |> 
               make.unique(sep = "_dup_")) %>%
  # lower case and remove white-space from rows
  mutate(across(where(is.character), \(x) str_squish(tolower(x)))) %>%
  # for stability mark other sex as NA since its only one case
  mutate(sex = ifelse(sex == "other", NA, sex),
         # re-calculate age variable (og is AAA)
         age = as.numeric(difftime(doa, dob, units = "days") / 365.25),
         sex = factor(sex, levels = c(0,1), labels = c("Female", "Male"))) %>% 
  # relocate age
  relocate(age, .after = doa) %>% 
  # rename(oc_adm = orbital_cellulitis_admissions,
  #        rv_ev = `rv/ev`) %>% 
  # remove last two rows containing summaries
  filter(!is.na(record_id)) %>% 
  # rename duplicate surgery variables
  rename(
    prior_surgery = surgery...28,
    surgery_adm = surgery...61) %>% 
  # convert variables to factors
  mutate(across(all_of(priori[priori != "age"]),
                ~ factor(.x, levels = c(0, 1), labels = c("No", "Yes"))),
         # collapse race/ethnicity into one column vector
         race_eth = case_when(
           hl == 1 ~ "Hispanic",
           nonhl == 1 & white == 1 ~ "White non-Hispanic",
           nonhl == 1 & black == 1 ~ "Black non-Hispanic",
           nonhl == 1 & `asian/pi`== 1 & native_american == 1 & other...10 == 1
           ~ "Other",
           TRUE ~ NA_character_))

# duplicate names
# which(names(oc_cleaned) == "surgery")
# names(oc_cleaned)[grepl("surgery", names(oc_cleaned))]

```


```{r}
# find duplicated column names
names(oc_raw)[duplicated(names(oc_raw))]
which(duplicated(names(oc_raw)))

# check where in my pipeline its creating a duplicate
n1 <- names(oc_raw)
n2 <- tolower(n1)
n3 <- gsub("\\s+", "_", n2)

n1[duplicated(n1)]
n2[duplicated(n2)]
n3[duplicated(n3)]
```

> The race/ethnicity covariate was categorized as White non-Hispanic, Black non-Hispanic, and Hispanic. We acknowledge that race is a social construct and that observed differences likely reflect structural inequities rather than biological differences.  

### Exploratory Data Analysis (EDA)


```{r label = "Used in final report as Table 1"}
tbl1 <- oc_cleaned %>%
  tbl_summary(
    by = surgery_adm,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, 
      spa, oa, ice, los),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "{mean} ({sd})",
        "{median} [{p25}, {p75}]",
        "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    #igits = list(AGE ~ 0, SYSBP ~ 0),
    missing = "ifany",
    missing_text = "NA",
    missing_stat = "{N_miss} ({p_miss}%)",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromise",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection`~ "NLDO",
      spa ~ "SPA",
      oa ~ "OA",
      ice ~ "ICE",
      los ~ "Length of Stay"
      )) %>%
  add_overall() %>% bold_labels() %>% italicize_levels() %>% add_n() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Surgery Needed**") %>%
  as_gt() %>%
  gt::fmt_markdown(columns = everything()) %>%
  tab_header(
    title = "Characteristics of Orbital Cellulitis Adult Cohort") %>% 
  tab_source_note(
    source_note = "Percentages are calculated within levels of the stratifying 
    variable") %>% 
  tab_options(
    table.width = pct(70),
    table.font.size = "small",
    heading.align = "left",
    data_row.padding = px(3),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

tbl1
```

For knitting to Word

```{r}
tbl1 <- oc_cleaned %>%
  tbl_summary(
    by = surgery_adm,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, 
      spa, oa, ice, los),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "Mean (SD)" = "{mean} ({sd})",
        "Median [Q1, Q3]" = "{median} [{p25}, {p75}]",
        "Min - Max" = "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "NA",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromised",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection` ~ "NLDO",
      spa ~ "SPA",
      oa ~ "OA",
      ice ~ "ICE",
      los ~ "Length of Stay")) %>%
  add_overall() %>%
  add_n() %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**Surgery Needed**") %>%
  modify_caption(
    "**Table 1: Characteristics of Orbital Cellulitis Adult Cohort**") %>%
  as_flex_table()

tbl1
```

```{r}
tbl2 <- oc_cleaned %>%
  tbl_summary(
    by = spa,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, 
      surgery_adm, oa, ice, los),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "Mean (SD)" = "{mean} ({sd})",
        "Median [Q1, Q3]" = "{median} [{p25}, {p75}]",
        "Min - Max" = "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "NA",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromised",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection` ~ "NLDO",
      surgery_adm ~ "Surgery During Admission",
      oa ~ "OA",
      ice ~ "ICE",
      los ~ "Length of Stay")) %>%
  add_overall() %>%
  add_n() %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**SPA**") %>%
  modify_caption(
    "**Table 1: Characteristics of Orbital Cellulitis Adult Cohort**") %>%
  as_flex_table()

tbl2
```

```{r}
tbl3 <- oc_cleaned %>%
  tbl_summary(
    by = oa,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, 
      surgery_adm, spa, ice, los),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "Mean (SD)" = "{mean} ({sd})",
        "Median [Q1, Q3]" = "{median} [{p25}, {p75}]",
        "Min - Max" = "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "NA",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromised",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection` ~ "NLDO",
      surgery_adm ~ "Surgery During Admission",
      spa ~ "SPA",
      ice ~ "ICE",
      los ~ "Length of Stay")) %>%
  add_overall() %>%
  add_n() %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**SPA**") %>%
  modify_caption(
    "**Table 1: Characteristics of Orbital Cellulitis Adult Cohort**") %>%
  as_flex_table()

tbl3
```


```{r}
tbl4 <- oc_cleaned %>%
  tbl_summary(
    by = ice,
    include = c(
      age, sex, race_eth, sinusitis, diabetes, immunocompromise, `drug/alc`,
      asthma, trauma, prior_surgery, dental_infection, `nldo/infection`, 
      surgery_adm, oa, spa, los),
    type = list(
      all_continuous() ~ "continuous2",
      all_categorical() ~ "categorical"),
    statistic = list(
      all_continuous() ~ c(
        "Mean (SD)" = "{mean} ({sd})",
        "Median [Q1, Q3]" = "{median} [{p25}, {p75}]",
        "Min - Max" = "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    missing = "ifany",
    missing_text = "NA",
    label = list(
      age ~ "Age",
      sex ~ "Sex",
      race_eth ~ "Race/Ethnicity",
      sinusitis ~ "Sinusitis",
      diabetes ~ "Diabetes",
      immunocompromise ~ "Immunocompromised",
      `drug/alc` ~ "Drug/Alcohol",
      asthma ~ "Asthma",
      trauma ~ "Trauma",
      prior_surgery ~ "Prior Surgery",
      dental_infection ~ "Dental",
      `nldo/infection` ~ "NLDO",
      surgery_adm ~ "Surgery During Admission",
      oa ~ "OA",
      spa ~ "SPA",
      los ~ "Length of Stay")) %>%
  add_overall() %>%
  add_n() %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(
    c("stat_1", "stat_2") ~ "**ICE**") %>%
  modify_caption(
    "**Table 1: Characteristics of Orbital Cellulitis Adult Cohort**") %>%
  as_flex_table()

tbl4
```


```{r}
# change label names for hospitals
unique(oc_cleaned$hospital)
```

Number of patients
```{r}
cat("Number of patients:",
  length(unique(oc_cleaned$record_id)),"\n\n")
```

Missing values
```{r}
oc_cleaned %>% ungroup() %>% miss_var_summary()
# missing values for predictors of interest
oc_cleaned[, priori] %>% ungroup() %>% miss_var_summary()
```


**NLDO** = Nasolacrimal duct infection:
a tear duct inflammation and/or infection, can happen in newborn infants and in
adults. Treatment usually includes antibiotics. Surgery is a possible option to 
stop recurrences.

**Dacryocystitis** is an infection of the tear sac (lacrimal sac) caused by a 
blockage in the tear duct (nasolacrimal duct), leading to pain, redness, and 
swelling around the eye. Treatment typically involves antibiotics, and in some 
cases, surgery may be necessary to relieve the blockage.

Neutrophils = innate, acute inflammation
Lymphocytes (T,B,NK) = adaptive, regulated immunity
  * Neutrophils ↑ = acute inflammation, stress, infection
  * Lymphocytes ↓ = stress response, immune suppression, chronic disease
  * So high NLR = More inflammation + relatively weaker adaptive immune response
  
Physiological meaning of high vs low NLR

High NLR:
  * Acute stress or infection
  * Chronic inflammation
  * Poor prognosis in some diseases

Low NLR:
  * Lower inflammation
  * Stronger relative adaptive immunity (context-dependent)

Get group counts
```{r}
count_grp_obs(oc_cleaned, vars = priori[priori != "age"], 
              mode = "separate")
```

```{r}
sapply(oc_cleaned[priori[priori != "age"]], table) 
```

Correlations
```{r label = "Correlations", fig.width=8, fig.height=6}
# need to convert back to numeric
X_bin <- oc_cleaned %>%
  dplyr::select(all_of(priori[priori != "age"])) %>%
  mutate(across(everything(), ~ as.numeric(.x) - 1))

clean_names <- c(
  "Sinusitis", "Diabetes", "Immunocompromise", "Drug/Alc",
  "Asthma", "Trauma", "Prior Surgery", "Surgery Needed",
  "Dental Infection", "NLDO Infection", "SPA", "OA", "ICE")

cor_matrix <- cor(X_bin, use = "pairwise.complete.obs", method = "pearson")
colnames(cor_matrix) <- clean_names
rownames(cor_matrix) <- clean_names

ggcorrplot(cor_matrix, type = "lower",lab = T)
```



Per Frank Harrell - Events per varialbe (EPV) ≈ 10 EPV rule
$EPV = \frac{\text{Events}}{\text{Predictors}}$

Variable Selection

```{r}
# form <- surgery_adm ~ .
# 
# step.fit <- stepwise(formula = form,
#                 data = oc_cleaned,
#                 type = "logit",
#                 include = c("age", "sex", "race_eth", "sinusitis", "diabetes",
#                             "immunocompromise", "drug/alc", "asthma", "trauma", 
#                             "prior_surgery", "dental_infection", "nldo/infection", 
#                             "spa", "oa", "ice"),
#                 strategy = "backward",
#                 metric = c("AICc", "SBC"))
# 
# summary(step.fit$backward$AICc)
# summary(step.fit$backward$SBC)
# performance(step.fit)
```


```{r}
# Check unique values and class of each predictor
vars_to_check <- c("age", "sex", "race_eth", "sinusitis", "diabetes",
                   "immunocompromise", "drug/alc", "asthma", "trauma", 
                   "prior_surgery", "dental_infection", "nldo/infection" 
                   ) #"spa", "oa", "ice"

for (v in vars_to_check) {
  if (v %in% names(oc_cleaned)) {
    cat(v, ":", 
        "class =", class(oc_cleaned[[v]]), 
        "unique values =", length(unique(na.omit(oc_cleaned[[v]]))), 
        "\n")
  } else {
    cat(v, "NOT FOUND in data\n")
  }
}
```

### Primary Analysis

#### Need for Surgery

**logistf**
logistic regression model using Firth's bias reduction method, equivalent to 
penalization of the log-likelihood by the Jeffreys prior
  * Profile likelihood = method used in statistics to estimate the likelihood 
  of a parameter by fixing other parameters and optimizing the remaining ones
  * Penalized likelihood = method that incorporates a penalty term into the 
  likelihood function to prevent overfitting and improve model estimation, 
  especially in cases with complex models or limited data.
  
> Variables were selected using penalized regression and then entered into a 
Firth logistic model for estimation.

> Candidate predictors were screened using cross-validated penalized logistic 
regression. Variables selected by glmnet were then entered into a Firth penalized 
logistic regression model along with prespecified adjustment covariates: age, 
sex, and race/ethnicity.

##### logistf

```{r}
vars_quoted <- ifelse(grepl("/", vars_to_check), 
                      paste0("`", vars_to_check, "`"), vars_to_check)
form <- as.formula(paste("surgery_adm ~", paste(vars_quoted, collapse = " + ")))
nfs.model.fbm <- logistf(form,data=oc_cleaned)

summary(nfs.model.fbm)
```

##### GLM
```{r}
nfs.model.glm <- glm(form,family="binomial",data=oc_cleaned)

summary(nfs.model.glm)
```

##### GLMnet

https://stackoverflow.com/questions/12937331/why-is-it-inadvisable-to-get-statistical-summary-information-for-regression-coef/17725220#17725220

"It is a very natural question to ask for standard errors of regression 
coefficients or other estimated quantities. In principle such standard errors 
can easily be calculated, e.g. using the bootstrap.

Still, this package deliberately does not provide them. The reason for this is 
that standard errors are not very meaningful for strongly biased estimates such 
as arise from penalized estimation methods. Penalized estimation is a procedure 
that reduces the variance of estimators by introducing substantial bias. The 
bias of each estimator is therefore a major component of its mean squared error,
whereas its variance may contribute only a small part.

Unfortunately, in most applications of penalized regression it is impossible to 
obtain a sufficiently precise estimate of the bias. Any bootstrap-based 
calculations can only give an assessment of the variance of the estimates. 
Reliable estimates of the bias are only available if reliable unbiased estimates
are available, which is typically not the case in situations in which penalized
estimates are used.

Reporting a standard error of a penalized estimate therefore tells only part of
the story. It can give a mistaken impression of great precision, completely 
ignoring the inaccuracy caused by the bias. It is certainly a mistake to make 
confidence statements that are only based on an assessment of the variance of 
the estimates, such as bootstrap-based confidence intervals do."
```{r}
# handles NA removal to be consistent with dimensions of outcome
mf <- model.frame(form, data = oc_cleaned)

X <- model.matrix(form, data = mf)[, -1] # remove y vector
y <- mf$surgery_adm
```

```{r}
# set so that covariates have a penalty factor of 0 so theyre always retained
penalty_factor <- rep(1, ncol(X))
penalty_factor[
  colnames(X) %in% c("age", "sexMale", "race_ethWhite non-Hispanic", 
                     "race_ethHispanic")] <- 0
cvfit <- cv.glmnet(
  X, y,
  family = "binomial",
  alpha = 1,
  type.measure = "class",
  penalty.factor = penalty_factor)

plot(cvfit)
```

extract best model
```{r}
#coef(cvfit, s = "lambda.min")
coef(cvfit, s = "lambda.1se")
# variables selected per cross-validation results
rownames(coef(cvfit, s = "lambda.1se"))[
  as.numeric(coef(cvfit, s = "lambda.1se")) != 0]
```

Predictions and model performance
```{r}
pred_prob <- predict(cvfit, newx = X, s = "lambda.min", type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

# ROC curve - evaluate PC1 as a predictor of malignancy on test set
roc_obj = roc(y, pred_prob)

# Find optimal sensitivity, specificity, and threshold using Youden's method
opt_coords <- coords(roc_obj, "best", 
                     ret = c("threshold", "sensitivity", "specificity"), 
                     best.method = "youden")

# Plot the ROC curve and display the AUC
plot(roc_obj, print.auc = TRUE, xlim = c(1, 0), ylim = c(0, 1))
text(0.4, 0.20, paste("Optimal Threshold:", round(opt_coords$threshold, 3)), 
     cex = 0.9, col = "steelblue")
text(0.4, 0.15, paste("Sensitivity:", round(opt_coords$sensitivity, 3)), 
     cex = 0.9, col = "steelblue")
text(0.4, 0.10, paste("Specificity:", round(opt_coords$specificity, 3)), 
     cex = 0.9, col = "steelblue")

auc(roc_obj)
```


#### Final Model

```{r}
# OR & 95% CI
fbm_tbl <- data.frame(
  #term = names(nfs.model.fbm$coefficients),
  OR = exp(nfs.model.fbm$coefficients),
  LL = exp(nfs.model.fbm$ci.lower),
  UL = exp(nfs.model.fbm$ci.upper),
  p_value = nfs.model.fbm$prob) %>% 
  mutate(
    # % change in odds (per unit increase)
    pct_chg = (OR  - 1) * 100,
    pct_LL = (LL - 1) * 100,
    pct_UL = (UL - 1) * 100) %>% 
  mutate(pct_ci = sprintf("(%.2f, %.2f)", pct_LL, pct_UL),
         OR_ci = sprintf("(%.2f, %.2f)", LL, UL)) 

fbm_tbl
```


```{r label = "Used in final report as table XX"}
nfs_tbl <- fbm_tbl %>%
  rownames_to_column("Predictor") %>% 
  mutate(n_fbm= nrow(model.frame(nfs.model.fbm)),
      Predictor = recode(
        Predictor,
        "age" = "Age (per 1yr)",
        "sexMale" = "Sex (Male vs Female)",
        "race_ethHispanic" = "HL vs Blk non-HL",
        "race_ethWhite non-Hispanic" = "White non-HL vs Blk non-HL)",
        "sinusitisYes" = "Sinusitis (Yes vs No)",
        "diabetesYes" = "Diabetes (Yes vs No)",
        "immunocompromiseYes" = "Immuno (Yes vs No)",
        "`drug/alc`Yes" = "Drugs/Alcohol (Yes vs No)",
        "asthmaYes" = "Asthma (Yes vs No)",
        "traumaYes" = "Trauma (Yes vs No)",
        "prior_surgeryYes" = "Prior Surgery (Yes vs No)",
        "dental_infectionYes" = "Dental (Yes vs No)",
        "`nldo/infection`Yes" = "NLDO (Yes vs No)",
        "spaYes" = "SPA (Yes vs No)",
        "oaYes" = "OA (Yes vs No)",
        "iceYes" = "ICE (Yes vs No)")) %>% 
  select(Predictor, n_fbm, OR, OR_ci, pct_chg, p_value) %>%
  filter(Predictor != "(Intercept)") %>% 
  gt() %>%
  cols_label(
    n_fbm = "N", 
    OR = "OR",
    pct_chg = "%Δ in Odds",
    OR_ci = "95% CI",
    p_value  = "P-value") %>%
  # tab_spanner(
  #   label = "Bayesian",
  #   columns = c(n_brm, bay_pct, bay_ci)) %>%
  fmt_number(columns = c(OR), decimals = 2) %>%
  fmt_number(columns = pct_chg, decimals = 1) %>% 
  fmt_number(columns = p_value, decimals = 3) %>%
    # bold specific text
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
    # highlight significant predictors <0.05
  tab_style(style = cell_fill(color = "#FFF3CD"),
            locations = list(
              cells_body(rows = p_value < 0.05),
              cells_stub(rows = p_value < 0.05))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = p_value),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "%Δ in Odds represents percent change in the odds of needing surgery
    in the predictor (computed as (OR − 1) × 100). Negative values indicate lower
    odds of surgical need.") %>% 
  tab_source_note(
    source_note = "Variables with very large odds ratios (e.g., SPA) reflect
    near-complete separation; effect sizes should be interpreted clinically 
    rather than numerically.") %>% 
    tab_header(
    title = "Factors Associated with Need for Surgery Among Adults Admitted for OC",
    subtitle = "Firth Logistic Regression Analysis") %>% 
  tab_options(
    table.font.size = "small",
    #heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

nfs_tbl
```
> SPA was strongly associated with surgery, with very high rates of surgery 
among affected patients

#### Length of Stay

```{r}
summary(oc_cleaned$los)
hist(oc_cleaned$los)
```


##### Negative Binomial

```{r}
los.nb.model <- glm.nb(
  los ~ age + sex + race_eth +
    sinusitis + diabetes + immunocompromise + `drug/alc` +
    asthma + trauma + prior_surgery + surgery_adm + dental_infection +
    `nldo/infection` , #spa + oa + ice
  data = oc_cleaned)

summary(los.nb.model)
```

```{r}
nb_tbl <- broom::tidy(los.nb.model, exponentiate = T, conf.int = T) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    pct_chg = (estimate  - 1) * 100,
    pct_LL = (conf.low - 1) * 100,
    pct_UL = (conf.high - 1) * 100) %>% 
  mutate(pct_ci = sprintf("(%.2f, %.2f)", pct_LL, pct_UL),
         RR_ci = sprintf("(%.2f, %.2f)", conf.low, conf.high))

nb_tbl
```


```{r}
los_tbl <- nb_tbl %>%
  mutate(n_los= nrow(model.frame(los.nb.model)),
      term = recode(
        term,
        "age" = "Age (per 1yr)",
        "sexMale" = "Sex (Male vs Female)",
        "race_ethHispanic" = "HL vs Blk non-HL",
        "race_ethWhite non-Hispanic" = "White non-HL vs Blk non-HL",
        "sinusitisYes" = "Sinusitis (Yes vs No)",
        "diabetesYes" = "Diabetes (Yes vs No)",
        "immunocompromiseYes" = "Immuno (Yes vs No)",
        "`drug/alc`Yes" = "Drugs/Alcohol (Yes vs No)",
        "asthmaYes" = "Asthma (Yes vs No)",
        "traumaYes" = "Trauma (Yes vs No)",
        "prior_surgeryYes" = "Prior Surgery (Yes vs No)",
        "surgery_admYes" = "Surgery During Admission (Yes vs No)",
        "dental_infectionYes" = "Dental (Yes vs No)",
        "`nldo/infection`Yes" = "NLDO (Yes vs No)",
        "spaYes" = "SPA (Yes vs No)",
        "oaYes" = "OA (Yes vs No)",
        "iceYes" = "ICE (Yes vs No)")) %>% 
  select(term, n_los, estimate, RR_ci, pct_chg, p.value) %>%
  filter(term != "(Intercept)") %>% 
  gt() %>%
  cols_label(
    term = "Predictor",
    n_los = "N", 
    estimate = "RR",
    pct_chg = "%Δ in LOS",
    RR_ci = "95% CI",
    p.value  = "P-value") %>%
  fmt_number(columns = c(estimate), decimals = 2) %>%
  fmt_number(columns = pct_chg, decimals =1) %>% 
  fmt_number(columns = p.value, decimals = 3) %>%
    # bold specific text
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  tab_style(
  style = cell_fill(color = "#FFF3CD"),
  locations = cells_body(rows = p.value < 0.05)) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = p.value),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "RR (rate ratio) represents the relative change in expected
    length of stay associated with each predictor. Values >1 indicate longer 
    LOS and values <1 indicate shorter LOS. Percent change in LOS is calculated
    as (RR − 1) × 100") %>% 
  tab_header(
    title = "Factors Associated with Length of Stay Among Adults Admitted for OC",
    subtitle = "Negative Binomial Regression Analysis") %>% 
  tab_options(
    table.font.size = "small",
    #heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

los_tbl
```

> Patients with XX had XX% longer/shorter expected length of stay (mean LOS), 
adjusting for other factors.


#### Subperiosteal abscess (SPA) 

```{r}
spa_vars <- c("age", "sex", "race_eth", "sinusitis", "diabetes",
              "immunocompromise", "drug/alc", "asthma", "trauma", 
              "prior_surgery", "surgery_adm","dental_infection",
              "nldo/infection") # "oa", "ice"

spa_vars_qt <- ifelse(grepl("/", spa_vars), 
                      paste0("`", spa_vars, "`"), spa_vars)

spa_form <- as.formula(paste("spa ~", paste(spa_vars_qt, collapse = " + ")))
spa.model <- logistf(spa_form, data=oc_cleaned)

summary(spa.model)
```


```{r}
# OR & 95% CI
fbm_spa_tbl <- data.frame(
  #term = names(nfs.model.fbm$coefficients),
  RR = exp(spa.model$coefficients),
  LL = exp(spa.model$ci.lower),
  UL = exp(spa.model$ci.upper),
  p_value = spa.model$prob) %>% 
  mutate(
    # % change in odds (per unit increase)
    pct_chg = (RR  - 1) * 100,
    pct_LL = (LL - 1) * 100,
    pct_UL = (UL - 1) * 100) %>% 
  mutate(pct_ci = sprintf("(%.2f, %.2f)", pct_LL, pct_UL),
         RR_ci = sprintf("(%.2f, %.2f)", LL, UL)) 

fbm_spa_tbl
```

```{r}
spa_tbl <- fbm_spa_tbl %>%
  rownames_to_column("Predictor") %>% 
  mutate(n_fbm= nrow(model.frame(spa.model)),
      Predictor = recode(
        Predictor,
        "age" = "Age (per 1yr)",
        "sexMale" = "Sex (Male vs Female)",
        "race_ethHispanic" = "HL vs Blk non-HL",
        "race_ethWhite non-Hispanic" = "White non-HL vs Blk non-HL)",
        "sinusitisYes" = "Sinusitis (Yes vs No)",
        "diabetesYes" = "Diabetes (Yes vs No)",
        "immunocompromiseYes" = "Immuno (Yes vs No)",
        "`drug/alc`Yes" = "Drugs/Alcohol (Yes vs No)",
        "asthmaYes" = "Asthma (Yes vs No)",
        "traumaYes" = "Trauma (Yes vs No)",
        "prior_surgeryYes" = "Prior Surgery (Yes vs No)",
        "surgery_admYes" = "Surgery in Admission (Yes vs No)",
        "dental_infectionYes" = "Dental (Yes vs No)",
        "`nldo/infection`Yes" = "NLDO (Yes vs No)",
        "oaYes" = "OA (Yes vs No)",
        "iceYes" = "ICE (Yes vs No)")) %>% 
  select(Predictor, n_fbm, RR, RR_ci, pct_chg, p_value) %>%
  filter(Predictor != "(Intercept)") %>% 
  gt() %>%
  cols_label(
    n_fbm = "N", 
    RR = "OR",
    pct_chg = "%Δ in Odds",
    RR_ci = "95% CI",
    p_value  = "P-value") %>%
  fmt_number(columns = c(RR), decimals = 2) %>%
  fmt_number(columns = c(pct_chg), decimals = 1) %>%
  fmt_number(columns = p_value, decimals = 3) %>%
    # bold specific text
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  # highlight significant predictors <0.05
  tab_style(style = cell_fill(color = "#FFF3CD"),
            locations = list(
              cells_body(rows = p_value < 0.05),
              cells_stub(rows = p_value < 0.05))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = p_value),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "%Δ in Odds represents percent change in the odds of needing surgery
    in the predictor (computed as (OR − 1) × 100). Negative values indicate lower
    odds of having an SPA.") %>% 
  tab_header(
    title = "Factors Associated with SPA Among Adults Admitted for OC",
    subtitle = "Firth Logistic Regression Analysis") %>% 
  tab_options(
    table.font.size = "small",
    #heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

spa_tbl
```


#### Orbital Abscess (OA) 

```{r}
oa_vars <- c("age", "sex", "race_eth", "sinusitis", "diabetes",
              "immunocompromise", "drug/alc", "asthma", "trauma", 
              "prior_surgery", "surgery_adm","dental_infection",
              "nldo/infection") #"spa", "ice"

oa_vars_qt <- ifelse(grepl("/", oa_vars), 
                      paste0("`", oa_vars, "`"), oa_vars)

oa_form <- as.formula(paste("oa ~", paste(oa_vars_qt, collapse = " + ")))
oa.model <- logistf(oa_form, data=oc_cleaned)

summary(oa.model)
```

```{r}
# OR & 95% CI
fbm_oa_tbl <- data.frame(
  #term = names(nfs.model.fbm$coefficients),
  RR = exp(oa.model$coefficients),
  LL = exp(oa.model$ci.lower),
  UL = exp(oa.model$ci.upper),
  p_value = oa.model$prob) %>% 
  mutate(
    # % change in odds (per unit increase)
    pct_chg = (RR  - 1) * 100,
    pct_LL = (LL - 1) * 100,
    pct_UL = (UL - 1) * 100) %>% 
  mutate(pct_ci = sprintf("(%.2f, %.2f)", pct_LL, pct_UL),
         RR_ci = sprintf("(%.2f, %.2f)", LL, UL)) 

fbm_oa_tbl
```

```{r}
oa_tbl <- fbm_oa_tbl %>%
  rownames_to_column("Predictor") %>% 
  mutate(n_fbm= nrow(model.frame(oa.model)),
      Predictor = recode(
        Predictor,
        "age" = "Age (per 1yr)",
        "sexMale" = "Sex (Male vs Female)",
        "race_ethHispanic" = "HL vs Blk non-HL",
        "race_ethWhite non-Hispanic" = "White non-HL vs Blk non-HL)",
        "sinusitisYes" = "Sinusitis (Yes vs No)",
        "diabetesYes" = "Diabetes (Yes vs No)",
        "immunocompromiseYes" = "Immuno (Yes vs No)",
        "`drug/alc`Yes" = "Drugs/Alcohol (Yes vs No)",
        "asthmaYes" = "Asthma (Yes vs No)",
        "traumaYes" = "Trauma (Yes vs No)",
        "prior_surgeryYes" = "Prior Surgery (Yes vs No)",
        "surgery_admYes" = "Surgery in Admission (Yes vs No)",
        "dental_infectionYes" = "Dental (Yes vs No)",
        "`nldo/infection`Yes" = "NLDO (Yes vs No)",
        "spaYes" = "SPA (Yes vs No)",
        "iceYes" = "ICE (Yes vs No)")) %>% 
  select(Predictor, n_fbm, RR, RR_ci, pct_chg, p_value) %>%
  filter(Predictor != "(Intercept)") %>% 
  gt() %>%
  cols_label(
    n_fbm = "N", 
    RR = "OR",
    pct_chg = "%Δ in Odds",
    RR_ci = "95% CI",
    p_value  = "P-value") %>%
  fmt_number(columns = c(RR), decimals = 2) %>%
   fmt_number(columns = c(pct_chg), decimals = 1) %>%
  fmt_number(columns = p_value, decimals = 3) %>%
    # bold specific text
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  # highlight significant predictors <0.05
  tab_style(style = cell_fill(color = "#FFF3CD"),
            locations = list(
              cells_body(rows = p_value < 0.05),
              cells_stub(rows = p_value < 0.05))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = p_value),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "%Δ in Odds represents percent change in the odds of needing surgery
    in the predictor (computed as (OR − 1) × 100). Negative values indicate lower
    odds of having an OA.") %>% 
  tab_header(
    title = "Factors Associated with OA Among Adults Admitted for OC",
    subtitle = "Firth Logistic Regression Analysis") %>% 
  tab_options(
    table.font.size = "small",
    #heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

oa_tbl
```


#### Intracranial Extension (ICE) 

```{r}
ice_vars <- c("age", "sex", "race_eth", "sinusitis", "diabetes",
              "immunocompromise", "drug/alc", "asthma", "trauma", 
              "prior_surgery", "surgery_adm","dental_infection",
              "nldo/infection") # "spa", "oa"

ice_vars_qt <- ifelse(grepl("/", ice_vars), 
                      paste0("`", ice_vars, "`"), ice_vars)

ice_form <- as.formula(paste("ice ~", paste(ice_vars_qt, collapse = " + ")))
ice.model <- logistf(ice_form, data=oc_cleaned)

summary(ice.model)
```


```{r}
# OR & 95% CI
fbm_ice_tbl <- data.frame(
  #term = names(nfs.model.fbm$coefficients),
  OR = exp(ice.model$coefficients),
  LL = exp(ice.model$ci.lower),
  UL = exp(ice.model$ci.upper),
  p_value = ice.model$prob) %>% 
  mutate(
    # % change in odds (per unit increase)
    pct_chg = (OR  - 1) * 100,
    pct_LL = (LL - 1) * 100,
    pct_UL = (UL - 1) * 100) %>% 
  mutate(pct_ci = sprintf("(%.2f, %.2f)", pct_LL, pct_UL),
         OR_ci = sprintf("(%.2f, %.2f)", LL, UL)) 

fbm_ice_tbl
```


```{r}
ice_tbl <- fbm_ice_tbl %>%
  rownames_to_column("Predictor") %>% 
  mutate(n_fbm= nrow(model.frame(ice.model)),
      Predictor = recode(
        Predictor,
        "age" = "Age (per 1yr)",
        "sexMale" = "Sex (Male vs Female)",
        "race_ethHispanic" = "HL vs Blk non-HL",
        "race_ethWhite non-Hispanic" = "White non-HL vs Blk non-HL)",
        "sinusitisYes" = "Sinusitis (Yes vs No)",
        "diabetesYes" = "Diabetes (Yes vs No)",
        "immunocompromiseYes" = "Immuno (Yes vs No)",
        "`drug/alc`Yes" = "Drugs/Alcohol (Yes vs No)",
        "asthmaYes" = "Asthma (Yes vs No)",
        "traumaYes" = "Trauma (Yes vs No)",
        "prior_surgeryYes" = "Prior Surgery (Yes vs No)",
        "surgery_admYes" = "Surgery in Admission (Yes vs No)",
        "dental_infectionYes" = "Dental (Yes vs No)",
        "`nldo/infection`Yes" = "NLDO (Yes vs No)",
        "spaYes" = "SPA (Yes vs No)",
        "oaYes" = "OA (Yes vs No)")) %>% 
  select(Predictor, n_fbm, RR, RR_ci, pct_chg, p_value) %>%
  filter(Predictor != "(Intercept)") %>% 
  gt() %>%
  cols_label(
    n_fbm = "N", 
    RR = "OR",
    pct_chg = "%Δ in Odds",
    RR_ci = "95% CI",
    p_value  = "P-value") %>%
  fmt_number(columns = c(RR), decimals = 2) %>%
  fmt_number(columns = c(pct_chg), decimals = 1) %>%
  fmt_number(columns = p_value, decimals = 3) %>%
    # bold specific text
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_stub(rows = everything())) %>%
  tab_style(style = cell_text(weight = "bold"),
    locations = cells_column_labels()) %>%
  # highlight significant predictors <0.05
  tab_style(style = cell_fill(color = "#FFF3CD"),
            locations = list(
              cells_body(rows = p_value < 0.05),
              cells_stub(rows = p_value < 0.05))) %>% 
  # p-value formatting like your other tables
  text_transform(
    locations = cells_body(columns = p_value),
    fn = function(x) ifelse(as.numeric(x) < 0.001, "<0.001",
                            format(round(as.numeric(x), 3), nsmall = 3))) %>%
  tab_options(table.font.size = "small") %>%
  tab_source_note(
    source_note = "%Δ in Odds represents percent change in the odds of needing surgery
    in the predictor (computed as (OR − 1) × 100). Negative values indicate lower
    odds of having an ICE.") %>% 
  tab_header(
    title = "Factors Associated with ICE Among Adults Admitted for OC",
    subtitle = "Firth Logistic Regression Analysis") %>% 
  tab_options(
    table.font.size = "small",
    #heading.align = "left",
    data_row.padding = px(3),
    table.width = pct(85),
    table.border.top.width = px(2),
    table.border.bottom.width = px(2))

ice_tbl
```







